---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Read in the zipped activity data included in the repository:

```{r, echo=TRUE}
data <- read.csv(unz("activity.zip", "activity.csv"))
```

The only data processing required is to 
```{r, echo=TRUE}
data$date <- as.Date(data$date)
```

The data is provided pre-processed and no further cleaning was applied.

## What is mean total number of steps taken per day?

The total number of steps taken per day is calculated by summing over all time intervals in a given day, ignoring `NA`'s in this case.

```{r, echo=TRUE}
library(plyr)
totsteps <- ddply(data, .(date), summarise, totSteps=sum(steps, na.rm = TRUE))
hist(totsteps$totSteps)
```

The mean total number of steps taken per day: 
```{r, echo=TRUE}
mean(totsteps$totSteps)
```

## What is the average daily activity pattern?

Take mean of the number of steps in a given time interval across all days included in the data set. Here, `NA`'s are ignored.

```{r, echo=TRUE}
avgSteps <- ddply(data, .(interval), summarise, avgSteps=mean(steps, na.rm = TRUE))
plot(avgSteps$interval, avgSteps$avgSteps, type="l")
```

The interval containing the maximum number of steps averaged across all days:
``` {r, echo=TRUE}
avgSteps$interval[which.max(avgSteps$avgSteps)]
```

## Imputing missing values

We investigate the possible bias in the above calculations introduced by the presence of missing values (`NA`).

First, we find that there are
```{r, echo=TRUE}
sum(is.na(data$steps))
```
step entries with missing values. This represents
```{r, echo=TRUE}
sum(is.na(data$steps)) / length(data$steps) * 100.
```
percent of the entries present in the data set.

To minimize this potential bias, `NA`'s are replaced with the average number of steps for the given interval.
```{r, echo=TRUE}
subNA <- merge(data, avgSteps)
subNA[is.na(subNA$steps), 2] <- subNA$avgSteps[is.na(subNA$steps)]
subNA <- subNA[,0:3]
```

Now recalculate the total number of steps per day
```{r, echo=TRUE}
totStepsNoNA <- ddply(subNA, .(date), summarise, totSteps=sum(steps, na.rm = TRUE))
hist(totStepsNoNA$totSteps)
```

The mean total number of steps taken per day: 
```{r, echo=TRUE}
mean(totStepsNoNA$totSteps)
```
This represents a 
```{r, echo=TRUE}
(mean(totsteps$totSteps) - mean(totStepsNoNA$totSteps)) / mean(totsteps$totSteps) * 100.
```
precent change compared to simply ignoring `NA`'s

The median total number of steps taken per day: 
```{r, echo=TRUE}
median(totStepsNoNA$totSteps)
```

## Are there differences in activity patterns between weekdays and weekends?
